---
title: "VITAL - Predicted Aminoacid Change - Data Analysis"
author: "André Soares"
date: "02/05/2020"
output: 
  html_document:
    theme: paper
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(readxl)
library(dplyr)
library(data.table)
library(UpSetR)
library(tidyverse)
```

```{r warning=FALSE, include=FALSE}
#list all VITAL folders
vital_dir = list.files("C:/Users/massa/Desktop/VITAL/DATA/WES Analysis/",
           pattern= "VITAL",
           full.names = T, include.dirs = T)

#extract only Predicted Aminoacid Change tables
for (i in vital_dir){
  tmp=list.files(i, 
             pattern = "Aminoacid",
             full.names = T)
  
  tmp_name = sapply(strsplit(i, "/"), tail, 1)
  #import ClinVar tables
  #name each with sample name
  for (a in tmp_name){
    df=read_excel(tmp)
    df$Sample=as.character(tmp_name)
    #call it by its right name
    assign(tmp_name, df)
    #keep a first row for the merge later
    if (a == "VITAL1"){
      first_row=head(df,1)
    }
    #clean up
    rm(df)
  }
}

#create master df
mVITAL=rbind(VITAL1,VITAL2,VITAL3,VITAL4,VITAL5,
              VITAL6,VITAL7,VITAL8,VITAL9,VITAL10,
              VITAL11,VITAL12,VITAL14,VITAL15,
              VITAL16,VITAL17,VITAL18,VITAL19,VITAL20,
              VITAL21,VITAL22,VITAL23,VITAL24,VITAL25,
              VITAL26,VITAL27,VITAL28,VITAL29)
```

```{r warning=FALSE, include=FALSE}
#turn Sample column into a binary matrix, keeping Phenotype and Gender
#retrieve frequency table
freq_smVITAL= mVITAL %>% 
  count(Sample, `Homo sapiens (hg38) _Gene`) %>%
  spread(Sample, n, fill=0)

#retrieve contingency table
abs_smVITAL= mVITAL %>% 
  count(Sample, `Homo sapiens (hg38) _Gene`) %>%
  spread(Sample, n, fill=0) %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>% 
  as.data.frame()

#turn new columns into integers (UpsetR-specific)
for(i in 2:ncol(abs_smVITAL)){
  abs_smVITAL[ , i] <- as.integer(abs_smVITAL[ , i]) 
}

#add metadata to master df
md=read_excel("C:/Users/massa/Desktop/VITAL/LVNC Samples Inventory_simplified.xlsx",
              sheet="Folha3")
mVITAL$Phenotype=md$Phenotype[match(mVITAL$Sample, md$`Project Code`)]
mVITAL$Gender=md$Gender[match(mVITAL$Sample, md$`Project Code`)]
```

```{r fig.width=10, include=FALSE}
#plot upsetr plot with metadata, save to image
# png("vital_upsetr.png", 
#     width = 3500, height = 2000,
#     res = 300)

upset(abs_smVITAL, nsets = 28, 
      order.by = "freq",
      nintersects = 200,
      set.metadata = list(data = md, 
                          plots = list(
                            list(
                              type = "matrix_rows", 
                              column = "Phenotype", 
                              assign = 10, 
                              colors = c(LVNC = "green", 
                                         Control = "navy",
                                         `LVNC/HCM` = "purple")),
                            list(
                              type = "heat",
                              column = "Gender",
                              assign = 10,
                              colors = c(M = "red",
                                         F = "yellow")
                            )
                          )),
      mb.ratio = c(.35, .65),
      )
# # 3. Close the file
# dev.off()
```

## Filtered data analysis

 - Creating an `UpSetR` plot to analyse any genes with 1 or more alleles (binary table) that are unique to a sample or shared between two or more.

 - Filtering out any alleles occurring in controls AND patients.

 - Removing alleles from from patient samples that appear in controls
 
 - Using column `Homo sapiens (hg38) _Gene`

 - Total number of altered genes excluding those present in controls:
```{r echo=FALSE}
#get vector list of `CLNDISDB clinvar_20191118.vcf` entries associated to Controls
ctrl_sv = mVITAL$`Homo sapiens (hg38) _Gene`[mVITAL$Phenotype == "Control"]
#remove rows 
f_mVITAL <- filter(mVITAL, !`Homo sapiens (hg38) _Gene` %in% ctrl_sv)

f_abs_smVITAL= f_mVITAL %>% 
  count(Sample, `Homo sapiens (hg38) _Gene`) %>%
  spread(Sample, n, fill=0) %>%
  mutate_if(is.numeric, ~1 * (. > 0)) %>% 
  as.data.frame()

f_freq_smVITAL= f_mVITAL %>% 
  count(Sample, `Homo sapiens (hg38) _Gene`) %>%
  spread(Sample, n, fill=0)

#unique genes NOT in controls, shared only by LVNC and LVNC/HCM
length(unique(f_mVITAL$`Homo sapiens (hg38) _Gene`))

#turn new columns into integers (UpsetR-specific)
for(i in 2:ncol(f_abs_smVITAL)){
  f_abs_smVITAL[ , i] <- as.integer(f_abs_smVITAL[ , i]) 
}
```

```{r echo=FALSE, fig.height=4, fig.width=7}
# png("f_vital_upsetr.png", 
#     width = 3500, height = 2000,
#     res = 300)

upset(f_abs_smVITAL, nsets = 28, 
      order.by = "freq",
      #43 found for > 1 intersection 
      nintersects = 43,
      set.metadata = list(data = md, 
                          plots = list(
                            list(
                              type = "matrix_rows", 
                              column = "Phenotype", 
                              assign = 10, 
                              colors = c(LVNC = "green", 
                                         `LVNC/HCM` = "purple")),
                            list(
                              type = "heat",
                              column = "Gender",
                              assign = 10,
                              colors = c(M = "red",
                                         F = "yellow")
                            )
                          )),
      mb.ratio = c(.35, .65),
      )
# # 3. Close the file
# dev.off()
```

 - Rows are samples (green for LVNC, purple for LVNC/HCM). Total number of altered genes per sample is plotted on the left vertical barchart.

 - Columns depict "interactions", i.e. altered genes unique to a sample or groups of genes shared by 2 or more samples. This is reflected on the vertical barchart on top.
 
 - Heatmap on the left illustrates the gender of patients.

### List of genes shared by more than one patient (control-removed):

Total number of genes in more than one sample:
```{r echo=FALSE}
f_abs_smVITAL_2 = as.data.frame(f_abs_smVITAL)
f_abs_smVITAL_2[2:16] <- lapply(f_abs_smVITAL_2[2:16], as.numeric)
f_abs_smVITAL_2$`Homo sapiens (hg38) _Gene`[is.na(f_abs_smVITAL_2$`Homo sapiens (hg38) _Gene`)] <- "no ID"
rownames(f_abs_smVITAL_2) <- f_abs_smVITAL_2[,1]
f_abs_smVITAL_2[,1] <- NULL

two_or_more_spl_list_f_abs_smVITAL = row.names(f_abs_smVITAL_2[rowSums(f_abs_smVITAL_2) > 1,])

two_or_more_spl_f_freq_smVITAL = filter(f_freq_smVITAL,
       `Homo sapiens (hg38) _Gene` %in% two_or_more_spl_list_f_abs_smVITAL)
length(two_or_more_spl_f_freq_smVITAL$`Homo sapiens (hg38) _Gene`)
```

Here's all of them!

```{r echo=FALSE, comment=NA}
print(unique(two_or_more_spl_f_freq_smVITAL$`Homo sapiens (hg38) _Gene`),
      row.names = FALSE, quote = F)
```

Table of variants per gene and per sample:

```{r}
library(kableExtra)
two_or_more_spl_f_freq_smVITAL %>%
  rename(Gene = `Homo sapiens (hg38) _Gene`) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = F)
```

